Deciding Bitvector Termination with Program Synthesis

### 2014-07-06 papers/termination/paper.tex

\begin{definition}[Non-Termination Formula -- Open Recurrence Set  {\bf [ONT]}]
\label{def:ont}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T_u(x, x') \wedge N(x') 
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula -- Closed Recurrence Set {\bf [CNT]}]
\label{def:cnt}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T_u(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

@@

Given that our program's state space is finite, a transition relation
induces a infinite execution if and only if some state is visited infinitely
often, or equivalently the state is visited twice in the course of the
execution.  This gives us an alternative way of expression a non-termination
proof: we can can try to find a witness $x$ to the satisfiability of the
formula
%
$$ \exists x . T^+(x, x)$$
%
However, deciding satisfiability of this formula directly requires a logic
that includes a transitive closure operator, which ours does not have.\\

\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based
on checking the satisfiability of formula {\bf [CNT]} may fail to find a
closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is
non-deterministic.  To understand the issue, imagine the following loop with
a non-deterministic assignment to \lstinline!y!:

\begin{lstlisting}[language=C]
while(x != 0) {
  y = nondet();
  x = x-y;
}
\end{lstlisting}

Only for certain choices of \texttt{y} the loop is non-terminating.  In
order to restrict the values that may be assigned to \texttt{y}, we need an
assumption associated with the non-deterministic choice.  However, such an
assumption may cause a reachable state at the non-deterministic assignment
to not have any successor.  In such a case the termination formula would be
satisfiable, although the execution halts.  This situation can be avoided by
checking the existence of at least one successor for each reachable state at
the non-deterministic assignment.  Although we did put a reasonable amount
of effort into it, we failed to come up with a formulation for the
non-termination of such non-deterministic transition relations inside the
synthesis fragment.
%
%formula {\bf [Nondet-CNT]} in Definition~\ref{def:nondet-cnt}.

\iffalse
\noindent {\bf Non-determinism.} A non-termination diagnosis procedure searching for a closed recurrence set under-approximates the transition relation, i.e. 
the closed recurrence set together with the under-approximation characterise non-termination \cite{DBLP:conf/tacas/ChenCFNO14}.
A potential difficulty with such a procedure lies in the handling of non-deterministic assignments. For illustration, consider the following loop:

\begin{lstlisting}[language=C]
while(x != 0) {
  y = nondet();
  x = x-y;
}
\end{lstlisting}

Only for certain values of $y$ the loop is non-terminating.
In order to restrict the values that may be assigned to $y$, we replace the non-deterministic choice with a fresh variable $a$ and use formula {\bf [CNT]}
to compute a closed recurrence set $N[a]$ parametrised by $a$. Now $N[a]$ becomes an assertion over $x \cup \{a\}$.
%, i.e. we under-approximate the transition relation. 
For this example, our technique computes $N[a](x,y,a) = ...$. \\%, where $a$ corresponds to the non-deterministic choice.  \\
%The issue with the closed recurrence set is that it yields an incomplete  for loops 
%that have non-deterministic assignments 
%parametric recurrence set

%% According to \cite{DBLP:conf/popl/GuptaHMRX08,DBLP:conf/tacas/ChenCFNO14}, a transition relation is non-terminating iff it has a recurrence set of states.
%% Candidate transition relations are 
%% \emph{lassos} in the program's CFG.
%% A lasso at a cutpoint location consists of two sequences
%% of transitions, the stem and the loop.
%% The stem is a finite path from an initial location to the cutpoint location. 
%% The loop is a finite path that starts and ends at the cutpoint
%% location by following a cyclic path through the CFG. 
%% A lasso induces an infinite execution if the infinite path obtained by traversing the stem and then unrolling
%% the loop infinitely many times is feasible. In this case, the lasso is said to be non-terminating.
%% We observe that in the case that $T$ is deterministic (that is, each state has
%% exactly one successor),  the non-termination problem is equivalent to the
%% satisfiability of Definition~\ref{def:deterministic-nonterm-formula}.

%% The method is incomplete as not all non-terminating
%% program executions are induced by lassos. An infinite behaviour
%% that is not periodic is illustrated by the following program:

%Our method is purely semantic.

@@

This issue does not arise when abstracting the environment as its under-approximation and the loop under consideration make use of existential quantification at the same nestedness level. However, when dealing with nested loops, we need to compute the under-approximation of the inner loop, which requires a nested existential quantifier, 
resulting in the $\exists \forall \exists$ alternation. 
%
%% computing the under-approximation of the loop's body requires an extra quantifier alternation. 
%% The origin of the problem comes from the fact that an under-approximation is an existential property and therefore
%% requires an extra existential quantifier, resulting in the $\exists \forall \exists$ alternation. This situation is similar to the one encountered when proving non-termination of non-deterministic loops. 
%
To address this limitation, we treat nested loops as a special case and give
a separate formula for them in Figure~\ref{fig:nonterm-nested}.  If the
formula on the right of the figure is satisfiable, then $L_1$ is
non-terminating as witnessed by the recurrence set $N_1$ and the initial
state $x_0$ in which the program begins executing.  There are two possible
scenarios for $L_2$'s termination:
%
\begin{itemize}

\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that
reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x)
\wedge P_2(x,x') \rightarrow N_1(x') $.

\item If $L_2$ is non-terminating, then $N_2$ is its recurrence set.

%% The issue appears when we try to under-approximate the body of the loop. 
%% Here we encounter a similar problem with the one for non-determinism mentioned above.

%% Consider the generic example in Figure~\ref{fig:nonterm-environment}, where we are attempting to prove non-termination
%% of the innermost loop $L_3$ by finding a closed recurrence set $N_3$. 
%% -- We will assume that the other two loops iterate only a finite number of times.

%% We now explain the constraints on the right of the figure.
%% Given that we prove non-termination, we try to  under-approximate the states from which $L_3$ can start executing. 
%% We do this in several steps:

%% $W_1$, $W_2$ and $W_3$ are inductive invariants corresponding to $L_1$, $L_2$ and $L_3$, respectively, such that:

%% - $W_1$ is an inductive invariant of $L_1$ that is strong enough to establish $W_2$ if $L_1$ exits (which it does by assumption).
%% - $W_2$ is an inductive invariant of $L_2$ that is strong enough to establish either $W_3$ or $N_3$ on entry to $L_3$, depending on whether $L_3$ terminates or not.
%% - Recall that $N_3$ is a recurrence set for $L_3$, so if a state $x \in N_3$ is reached, the loop $L_3$ will execute forever. Otherwise, $W_3$ holds on entry to $L_3$.
%% - $W_3$ is an inductive invariant of $L_3$ that is strong enough to re-establish $W_2$ when $L_3$ exits.

%% If the non-termination formula is satisfiable, $x_0$ is some initial state in which the program begins executing.

%% First, we under-approximate the states in which $L_2$ can start executing through $W_2$.
%% Then $N_1$ is an inductive invariant of $L_1$ that is strong
%% enough to establish $N_2$ if $L_1$ exits (which it does by assumption).  Now, $N_2$ is an inductive invariant
%% of $L_2$ that is strong enough to establish either $N_3$ or $N_4$ on entry to $L_3$.  Recall that $N_4$ is
%% a recurrence set for $L_3$, so if a state $x \in N_4$ is reached, the loop $L_3$ will execute forever.
%% Otherwise, $N_3$ holds on entry to $L_3$.  $N_3$ is an inductive invariant of $L_3$ that is strong enough to
%% re-establish $N_2$ when $L_3$ exits.

%% If the constraint system is satisfiable, $x_0$ is some initial
%% state in which the program begins executing.

%%   Since we assumed that $L_2$ iterated only a finite number of times,
%% $N_3$ must be established only a finite number of times, which means that $N_4$ will eventually be reached,
%% at which point $L_3$ will loop forever since $N_4$ is a recurrence set for $L_3$.

%% \begin{figure*}
%% \begin{framed}
%%  \begin{minipage}{0.19\textwidth}
%% \begin{lstlisting}
%% L1: while (c1) {
%%      B1;
%%     }

%% L2: while (c2) {
%%      P;

%% L3:  while (c3) {
%%        B3;
%%      }
%%     }
%% \end{lstlisting}
%% \end{minipage}
%% \vline
%% \begin{minipage}{0.82\textwidth}
%% \begin{align*}
%%  \exists W_1, W_2, W_3, N_3, x_0 . \forall x, x' . \\
%%   W_1(x_0) \, \wedge \\
%%   c_1(x) \wedge W_1(x) \wedge B_1(x, x') & \rightarrow W_1(x') \, \wedge \\
%%   \lnot c_1(x) \wedge W_1(x) & \rightarrow W_2(x) \, \wedge \\
%%   W_2(x) & \rightarrow c_2(x) \, \wedge \\
%%   W_2(x) \wedge P(x, x') & \rightarrow W_3(x') \vee N_3(x') \, \wedge \\
%%   c_3(x) \wedge W_3(x) \wedge B_3(x, x') & \rightarrow W_3(x') \, \wedge \\
%%   \lnot c_3(x) \wedge W_3(x) & \rightarrow W_2(x) \, \wedge \\
%%   N_3(x) & \rightarrow c_3(x) \, \wedge \\
%%   N_3(x) \wedge B_3(x, x') & \rightarrow N_3(x')
%% \end{align*}
%% \end{minipage}
%% \end{framed}

%% \caption{The constraint system for non-termination\label{fig:nonterm-environment}}
%% \end{figure*}

>>>>

\begin{definition}[Non-Termination Formula -- Open Recurrence Set  {\bf [ONT]}]
\label{def:ont}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~\wedge \\ &  N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T_u(x, x') \wedge N(x') 
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula -- Closed Recurrence Set {\bf [CNT]}]
\label{def:cnt}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge \\ & N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T_u(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

@@

Given that our program's state space is finite, a transition relation
induces a infinite execution if and only if some state is visited infinitely
often, or equivalently the state is visited twice in the course of the
execution.  This gives us an alternative way of expression a non-termination
proof: we can can try to find a witness $x$ to the satisfiability of the
formula
%
$$ \exists x . T^+(x, x)$$
%
However, deciding satisfiability of this formula directly requires a logic
that includes a transitive closure operator, which ours does not have.

@@

This issue does not arise when abstracting the environment as its under-approximation and the loop under consideration make use of existential quantification at the same nestedness level.

However, when dealing with nested loops, we need to  under-approximate the inner loop, requiring a nested existential quantifier, 
resulting in the $\exists \forall \exists$ alternation.  At first sight, this seems fatal for our formulation.  Luckily, it is possible to
reformulate non-termination as an $\exists \forall$ formula by observing that, unlike a ranking function, the defining property of a
recurrence set is \emph{non relational} -- if we end up in the recurrence set, we don't care exactly where we came from as long as
we know that it was also somewhere in the recurrence set.  This allows us to cast non-termination of nested loops as the formula
shown in Figure~\ref{fig:nonterm-nested}.

If the formula on the right of the figure is satisfiable, then $L_1$ is non-terminating as witnessed by the recurrence set $N_1$ and
the initial state $x_0$ in which the program begins executing. There are two possible scenarios for $L_2$'s termination:
\begin{itemize}
\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') \rightarrow N_1(x') $. 
\item If $L_2$ is non-terminating, then $N_2 \wedge G_2$ is its recurrence set.
\end{itemize}
